---
title: "Model Predicted Enviromental Suitability for Dengue"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Predicted Enviromental Suitability for Dengue}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Sensitivity between LighGBM and XGBoost


Para todos los periodos de tiempo analizados, el algoritmos de Machine Learning LightGBM presentó sensibilidades mayores y diferencias menores entre el train y test comparado con el algoritmos de Machine Learning XGBoost. 

```{r sensibilidad, fig.cap= "Figura 1. Sensibilidad en el train y test con LightGBM y XGBoost"}
#| warning: False
#| message: False
#| echo: true
#| output: true
#| out.width: "100%"
#| dpi: 600
# Step 1. load the datasets ####
load("~/Library/CloudStorage/Dropbox/projects/INSP/spatial_temporal_distribution_dengue_mx/9.output/2.results_ml/metrics/metrics.RData")


# Step 2. visualization os sensitivity ####
metrics |>
    dplyr::distinct() |>
    dplyr::filter(.metric %in% c("sens"))  |>
    ggplot2::ggplot(ggplot2::aes(y = .train,
                                 x = .test,
                                 label = period,
                                 fill = lm)) +
    ggplot2::geom_point(shape = 21,
                        size = 5,
                        #fill = "black",
                        color = "white") +
    ggplot2::scale_fill_manual("ML",
                               values = c("#E01A59", "#63C1A0")) +
    ggplot2::geom_text(size = 1.5,
                       vjust = -1) +
    ggplot2::coord_cartesian(xlim = c(0.2, 1),
                             ylim = c(0.2, 1)) +
    ggplot2::scale_x_continuous(breaks = seq(from = 0.2, 
                                             to = 1, 
                                             by = 0.2)) +
    ggplot2::scale_y_continuous(breaks = seq(from = 0.2, 
                                             to = 1,
                                             by = 0.2)) +
    ggplot2::ylab("Sensibilidad (Train)") +
    ggplot2::xlab("Sensiblidad  (Test)") +
    ggplot2::geom_abline(intercept = 0, 
                         slope = 1, 
                         color="gray60", 
                         linetype= 1, 
                         alpha = 0.5,
                         linewidth=1.5) +
    ggplot2::theme(panel.border = ggplot2::element_rect(colour = "black",
                                               fill = NA, 
                                               linewidth = 1),
                   legend.position = "inside",
                   legend.position.inside = c(0.9, 0.15),
                   legend.background = ggplot2::element_blank())
```



## Model Predicted Enviromental Suitability for Dengue for 2008-2009 dataset

```{r hs_2008_2009, fig.cap= "Figura 2. Model Predicted Enviromental Suitability for Dengue for 2008-2009 dataset"}
library(terra)
library(tidyterra)
library(sf)
library(ggplot2)
# Step 1. load the raster layers of the probabilities ####
terra::rast("/Users/fdzul/Library/CloudStorage/Dropbox/projects/INSP/spatial_temporal_distribution_dengue_mx/9.output/2.results_ml/den2008_2009_lightgbm_60K.tif") |>
    dplyr::select(.pred_presence) |>
mapgl::maplibre_view(palette = viridis::turbo) |>
    mapgl::add_continuous_legend(legend_title = "Probabilidad (2008-2009)",
                                 position = "bottom-left",
                                 values = c(0, 0.25, 0.50, 0.75, 1),
                                 colors = viridis::turbo(n = 10)) |>
    mapgl::add_fullscreen_control(position = "top-left") |> 
    mapgl::add_navigation_control() |>
    mapgl::add_globe_control() |>
    mapgl::add_scale_control()

```


## Model Predicted Enviromental suitability for Dengue for 2012-2013 dataset

```{r hs_2012_2013, fig.cap= "Figura 3. Model Predicted Enviromental Suitability for Dengue for 2012-2013 dataset"}
# Step 1. load the pkgs ####
library(terra)
library(tidyterra)
library(sf)
library(ggplot2)
# Step 1. load the raster layers of the probabilities ####
terra::rast("/Users/fdzul/Library/CloudStorage/Dropbox/projects/INSP/spatial_temporal_distribution_dengue_mx/9.output/2.results_ml/den2012_2013_lightgbm_60K.tif") |>
    dplyr::select(.pred_presence) |>
mapgl::maplibre_view(palette = viridis::turbo) |>
    mapgl::add_continuous_legend(legend_title = "Probabilidad (2012-2013)",
                                 position = "bottom-left",
                                 values = c(0, 0.25, 0.50, 0.75, 1),
                                 colors = viridis::turbo(n = 10)) |>
    mapgl::add_fullscreen_control(position = "top-left") |> 
    mapgl::add_navigation_control() |>
    mapgl::add_globe_control() |>
    mapgl::add_scale_control()
```

## Model Predicted Enviromental suitability for Dengue for 2018-2020 dataset

```{r hs_2018_2020, fig.cap= "Figura 4. Model Predicted Enviromental Suitability for Dengue for 2018-2020 dataset"}
# Step 1. load the pkgs ####
library(terra)
library(tidyterra)
library(sf)
library(ggplot2)
# Step 1. load the raster layers of the probabilities ####
terra::rast("/Users/fdzul/Library/CloudStorage/Dropbox/projects/INSP/spatial_temporal_distribution_dengue_mx/9.output/2.results_ml/den2018_2020_lightgbm_60K.tif") |>
    dplyr::select(.pred_presence) |>
mapgl::maplibre_view(palette = viridis::turbo) |>
    mapgl::add_continuous_legend(legend_title = "Probabilidad (2018-2020)",
                                 position = "bottom-left",
                                 values = c(0, 0.25, 0.50, 0.75, 1),
                                 colors = viridis::turbo(n = 10)) |>
    mapgl::add_fullscreen_control(position = "top-left") |> 
    mapgl::add_navigation_control() |>
    mapgl::add_globe_control() |>
    mapgl::add_scale_control()
```

## Model Predicted Enviromental suitability for Dengue for 2023-2025 dataset

```{r hs_2023_2025, fig.cap= "Figura 5. Model Predicted Enviromental Suitability for Dengue for 2023-2025 dataset"}
# Step 1. load the pkgs ####
library(terra)
library(tidyterra)
library(sf)
library(ggplot2)
# Step 1. load the raster layers of the probabilities ####
terra::rast("/Users/fdzul/Library/CloudStorage/Dropbox/projects/INSP/spatial_temporal_distribution_dengue_mx/9.output/2.results_ml/den2023_2025_lightgbm_60K.tif") |>
    dplyr::select(.pred_presence) |>
mapgl::maplibre_view(palette = viridis::turbo) |>
    mapgl::add_continuous_legend(legend_title = "Probabilidad (2023-2025)",
                                 position = "bottom-left",
                                 values = c(0, 0.25, 0.50, 0.75, 1),
                                 colors = viridis::turbo(n = 10)) |>
    mapgl::add_fullscreen_control(position = "top-left") |> 
    mapgl::add_navigation_control() |>
    mapgl::add_globe_control() |>
    mapgl::add_scale_control()
```

